Data-Driven Modelling \& \\ Forecasting of Chaotic Systems

### 2022-01-13 Gandhi_Project_Report/draft3.tex
Nearly done with changes

\title{Project Report}

\author{} 
\date{}
\date{}
\maketitle
\tableofcontents

@@

The more intricate details of proofs are referred to where relevant throughout. This report is organised as follows: 
\newline In Chapter~\ref{ch2}, we recall the definition of a discrete-time dynamical system, how a discrete-system arises from a flow of an ODE and then proceed to define the inverse limit space and topological conjugacy of autonomous systems. 
\newline In Chapter~\ref{ch3}, we introduce the problem of forecasting dynamical systems, state the Takens delay embedding theorem and discuss various issues faced while forecasting. 
\newline In Chapter~\ref{ch4}, a driven dynamical system is defined and discuss the properties of a specific class of driven dynamical systems that we make use of in this thesis/report/paper. 
\newline Finally in Chapter~\ref{ch5} we show the implementation of these forecasting methods, and conclusions are provided in Chapter~\ref{ch6}.

@@

\section{Taken Embedding Theorem}

@@

@@

\begin{Definition}
  [\bf Entire Solution] \label{Dfn_Soln} \rm
  A sequence $\{x_n\}_{n\in\mathbb{Z}}\subset X$ is called an entire solution(or simply a solution) of the driven system  $g$ with input $\overline{u}$ when it satisfies 
  \[g(u_{n-1}, x_{n-1})=x_n\] for all $n\in\mathbb{Z}$
\end{Definition}

@@

\begin{Definition}
  [\bf Unique Solution Property] \label{Dfn_usp}\rm
  A driven system $g$ is said to have the Unique Solution Property (USP) if for each input $\overline{u}$ there exists exactly one solution. 
  Alternatively we may formulate the USP as follows: $g$ has the Unique Solution Property if there exists a well-defined map $\Psi:{U}\to{X}$ with $\Psi({\overline{u}})$ denoting the unique solution.
\end{Definition}

One of the first result obtained upon defining the USP, is the fact that every solution will attract any different initial conditions towards the component parts of the solution. %, i.e. that the solution $\Psi$ is a non-autonomous uniform attractor. \ednote{Define Uniform Attractor before stating theorem or else omit the statement.} 
If $g$ has the USP, then any solution to $g$ is a uniform attractor in the nonautonomous dynamical systems literature \cite{Manju_Nonlinearity}. The discussion on nonautonomous attractors is beyond the scope of this thesis and we refer the reader to \cite{Manju_ESP, esann2012ids} for further reading.

% \ednote{M: You can omit the theorem, and just say it is a uniform attractor in the nonautonomous dynamical systems literature \cite{Manju_Nonlinearity}, and say the discussion on nonautonomous attractors is beyond the scope of the thesis.}
% Paraphrased.
 In our context, $g$ having the USP is equivalent to $g$ being a topological contraction \cite{manjunath2021universal}, or alternatively that $g$ exhibits the Echo State Property (ESP).

This property is used to describe the concept of fading memory or memory-loss in the majority of literature within the field reservoir computing (RC) \textbf{(Citation)}. 
$g$ having the ESP guarantees that the entire left-infinite history of an input will determine exactly the current system state; there is only ever the possibility of a single reachable state at a given instance of time \cite{jaeger2001echo}
The ESP is also often discussed in terms of a stability property that holds for all inputs of a system~\cite{manjunath2020stability} and, furthermore, plays a key role in the design and training of recurrent neural networks within the field of RC \cite{Manju_ESP}.

Defined formally, the ESP is intrinsically linked to an input (see \cite{Manju_ESP}) Unfortunately, definitions and results concerning the ESP do not have much practical use since a relationship of the network's response to the temporo-statistical properties of the input is not directly described. A more detailed presentation may be found in~\cite{jaeger2001echo}.

@@

Having already defined the reachable set $X_U$, we now pause for a moment to fix additional notation.
Letting $\cev{u}^{n}:=(\ldots,u_{n-2} ,u_{n-1})$ be the left-infinite subsequence of an input, $\overleftarrow{U}$ then denotes all these left-infinite sequences in $U$. 
Moreover, $\cev{u}^{n}v:=(\ldots,u_{n-2} ,u_{n-1}, v)$ is to symbolise the input up to time $n$ with $v \in U$ being the specific input value at time $n$. 
The introduction of a new input at time $n$ can be described by a mapping $\sigma_v:   \cev{u}^{n} \mapsto \cev{u}^{n}v$.

The question now becomes whether we may establish a semi-conjugacy %\ednote{M: I have changed it to a semi-conjugacy since conjugacy amounts to asking a infinite-dimensional space $\overleftarrow{U}$ to be embedded in $X_U$ which is not possible.} 
as presented below for the driven system $g$.

@@

Continuing to refine our attention, we proceed to consider a specific subclass of conjugacies.

  \begin{Definition} 
    [\bf Universal Semi-Conjugacy]\label{Def_UnivSemiConj}
    Given a driven system $g$, we  call a continuous and surjective map $h : \overleftarrow{U} \to X_U$ a universal semi-conjugacy if  diagram \ref{Scomm_h} commutes for all $v \in U$.
  \end{Definition}

If the universal semi-conjugacy $h$ exists (i.e. the diagram in \ref{Scomm_h} commutes) then the solution $\Psi$ will be at least a coarse-grained representation of the input $u$.

Does such a function $h$ for the driven system defined above exist? If $g$ has the USP and $\Psi(u)=\{x_n\}_{n\in\mathbb{Z}}$ then $h$, defined by  $h(\cev{u}_n):=g(u_n,x_{n-1})=x_n$, will satisfy the semi-conjugacy in the graph above \ref{Scomm_h}.
Regrettably, such a continuous mapping $h$ is not guaranteed to exist \cite{M: Have edited here} when $g$ does not have the USP \cite[Lemma 5]{Manju_Nonlinearity}.
Note that even when $h$ does exist, we are not guaranteed its injectivity. Considering again example \ref{ex_halfux}, we see that even if $h$ were to exist, it could not be injective as $X_U=\{0\}$.

@@

\begin{Definition}
  [\bf Causal Mapping]\label{Def_CausMap}
  A continuous, surjective map $H:\overleftarrow{U}\to\overleftarrow{X}_U$ such that \[H\circ\tilde{g}_v=\sigma\circ{H}\] where $\tilde{g}_v$ maps $(\ldots, u_{-2}, u_{-1})$ to $(\ldots, u_{-2}, u_{-1}, g(v, u_{-1}))$.
\end{Definition}

@@

 \begin{Theorem}
  For a compactly driven system, a causal mapping $H$ exists if and only if $g$ has the USP. 
\end{Theorem}
\begin{proof}
  See~\cite[Th.3]{manjunath2013echo}.
\end{proof}

We direct our attention to the function \[H_2{\overline{}} := (h(r\overline{u}, h(\overline{u}))) \]

An embedding of the space $\overleftarrow{U}$  in $\overleftarrow{U}$ would allow one to establish topological conjugacy, which in turn provides stronger results than merely obtaining a coarse-grained representation via a causal mapping \cite{M: I feel this is very confusing to the reader as the co-domain of causal mapping is the left-infinite space and , so please omit if possible. You can just say we want to restrict inputs to the left-infinite orbits of a dynamical system and then explore conjugacy instead of a mere semi-conjugacy}.

Before formalising this, we make mention of the concept of the inverse-limit system of a dynamical system $(U,T)$. 
These spaces are noteworthy as their topology can help one ascertain much about the original system \cite{Manju_IEEE}. The topological structure for inverse-limit systems of chaotic dynamical systems is complicated, so we relegate further discussion to \cite{kennedy2008inverse_limit, ingram2011inverse}.
In broad sweeps, the inverse-limit space of a dynamical system is a self-map on a subset of an infinite-dimensional space where each point in the inverse-limit space corresponds to a backward orbit of the map $T$. 
Following \cite{manjunath2021universal, ingram2011inverse}:

@@

One must be careful to avoid a choice of $g$ which would  add complexity to the obtained solution.

@@

\begin{Definition}
  [\bf SI-Invertibility]\label{Dfn_SIinv}\rm
  A map $g$ is said to be SI-Invertible if $g$ is invertible for all $x\in{X}$. Alternatively it may be said that if, given $x_n$ and $x_{n-1}$, $u_n$ can be uniquely determined from $x_n=g(u_n,x_{n-1})$, then $g$ is said to be SI-invertible.
\end{Definition}

@@

\begin{equation}  \label{eqn_driving}
  g(u,x) = (1-a)x + a\overline{\tanh}(Au + \alpha Bx)
\end{equation} 
is both SI-invertible and, if $\alpha\beta<1$, also possesses the USP. 
One may easily show that $g$ is SI-invertible by recovering $u_n$ by 
\begin{equation} \label{eqn_SI_RNN}
  u_n := A^{-1}\bigg(\overline{\tanh}^{^{-1}}\frac{1}{a}\Big(x_{n+1}-(1-a)x_n\Big) \bigg) - \alpha B x_n
  \end{equation}
  for $x_{n-1}, x_n$ known.

@@

\begin{equation} \label{Fig_inverse_limsystem}
  %  \[ 
      \psset{arrows=->, arrowinset=0.25, linewidth=0.6pt, nodesep=3pt, labelsep=2pt, rowsep=0.7cm, colsep = 1.1cm, shortput =tablr}
      \everypsbox{\scriptstyle}
      \begin{psmatrix}
      \widehat{U}_T  && \widehat{U}_T \\
      Y_T && Y_T \\
      %%%
     %  \ncline{1,1}{1,2}^{\widehat{T}} \ncline{1,1}{2,1} <{h}
     %  \ncline{1,2}{2,2} > {H_2}
     %  \ncline{2,1}{2,2}^{G_T}
      \end{psmatrix}
    %  \]
  \end{equation}

@@

The graph in equation \ref{Fig_inverse_limsystem} is then extended as below:

@@

From \eqref{eqn_driving} we know that $ g(u,x)=(1-a)x+a{\cdot}\tanh(Au+\alpha{a}Bx)$. Recall that $g$ is SI-invertible. %Adopting the terminology of \cite{manjunath2021universal}, $A$ is called the “input matrix” and $B$ the “reservoir matrix”.

We make use of a Recurrent Neural Network (RNN) in our implementation. RNN's are widely used in the literature for discrete-time processing and are considered to be discrete-time version of continuous dynamical systems modelling a network of neurons. (\textbf{Citation.})
A complete discussion of RNN's is outside the scope of this thesis, so we present only the details necessary to the current discussion. For information, (\textbf{Cite, Cite, Cite}) may be consulted. We implement our RNN in the \emph{Python} programming language and making use of the \emph{Scipy} and \emph{Tensorflow} libraries.

@@

We made several changes which gave improved results:
\vspace{-8mm}
\begin{enumerate}
\item Using Delay Coordinates (DC) when driving the system vs. using Single/Multi Delay Dynamics (SDD/MDD) after driving (and before training). Explain/show why these two methods are equivalent. \ednote{Would MDD  be not computationally intensive ? So may be you just explain if you can’t show it.}
\item Sparse matrix implementation. Slight improvement in some scenarios, but not fantastic. \ednote{M: Did it not help you a great deal with the DP forecasting? B: Sparse matrices tended to be finicky. In some cases they helped whereas in others they actually worsened the solution somewhat. There's some tuning to be done in terms of condition number, scaling and density}

(Possible paragraph here/later on implementing sparsity of A/B: Generate sparse matrix that is also well-conditioned and correctly scaled matrix.)
\item Density of matrices A/B(i.e. number of non-zero entries in each matrix) (separate from sparse matrix implementation)
\end{enumerate}

@@

The scalar equations of motion are derived using Newtonian physics and can be found in~\cite{DPFormulas}. The second-order system is converted to a first-order ODE and solved using the Runge-Kutta numerical method.  Noise-free, undamped DP data-sets were generated and utilised in this project due to the requirement that the map $T$ be surjective.
It may easily be verified that a DP undergoing damping (as seen in \ref{conjugacy_with_Gamma}) is not surjective and hence the required assumptions to apply our theory do not hold. \ednote{B: Change sentence.}

\begin{figure}[ht]
  \includegraphics[scale=0.4]{_dp_dying.eps}
  \centering
\caption{A damped pendulum is not surjective.}
\label{fig:damped_pendulum}
\end{figure}

\textbf{Parameter Values \& Successes}

Train length: 15000. Prediction Length: 1000. Discard to wait for network to forget: 500. Dimension of network: 128 neurons.
$\alpha=0.99$. $a=0.5$.

Hidden Layers: 16. Layer dimension: 64. Epochs:256. Batch Size: 128. Delay(for MDD): 8. Density of Sparse Matrices: 0.2. $d=1$

\ednote{B: Write this paragraph.}

\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{_DP_Fig-2F2023.eps} 
  \captionof{figure}{Predicted(red) and actual(blue) trajectories of the $x$- and $y$-coordinates of each of the pendulum heads.} 
\end{figure}
\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{_DP_Fig-2B2023.eps}
  \captionof{figure}{Densities of predicted and actual values of x-coordinate of first pendulum head. }
  \end{figure}
  \begin{figure}[ht]
    \centering 
  \includegraphics[width=\linewidth]{_DP_Fig-2D_5.eps}
    \captionof{figure}{Takens Delay Map with delay=5 to illustrate behaviour of predicted and actual values of double pendulum data.}
\end{figure}

\subsection{Other attractors I’ve experimented with: Clifford \& Thomas}
We follow the methodology presented in \cite{manjunath2021universal} but opt to consider attractors different from the Lorenz system and H\'enon map in this project.

@@

The behaviour of the system with a $\beta$-parameter value of 0.1056 is simulated by scaling it to fit inside the interval $[-1,1]$ and both perturbed with noise as well as leaving it intact. Specifically the noise is obtained from a normally distribution with mean 0 and standard deviation 0.05, which translates approximately to *dB.
 A sequence of 3000 observations in the clear and noisy data-set respectively is fed into the system with 300 entries discarded before predicting the next 9000 entries.

@@

The RNN was initialised as follows: The dimension of the network is set equal to 500 neurons and 12 hidden layers of dimension 64 each were initialised. Training is realised through 150 epochs of batch sizes of 128 each; the map is learnt via the Adam Optimiser using finer and finer learning rates (0.001 and then divided by 10 at every iteration).

@@

\ednote{B: Expand}

>>>>

\title{Data-Driven Modelling \& \\ Forecasting of Chaotic Systems}

\author{B.L. Nortier} 
\date{January, 2022}
\date{}
\maketitle
\tableofcontents

% \begin{abstract}
%   Abstract goes here.
%   \end{abstract}

@@

The more intricate details of proofs are referred to where relevant throughout. This report is organised as follows: 
\newline In Chapter~\ref{ch2}, we recall the definition of a discrete-time dynamical system, how a discrete-system arises from a flow of an ODE and then proceed to define the inverse-limit space and topological conjugacy of autonomous systems. 
\newline In Chapter~\ref{ch3}, we introduce the problem of forecasting dynamical systems, state the Takens delay embedding theorem and discuss various issues faced while forecasting. 
\newline In Chapter~\ref{ch4}, a driven dynamical system is defined and discuss the properties of a specific class of driven dynamical systems that we make use of in this thesis/report/paper. 
\newline Finally in Chapter~\ref{ch5} we show the implementation of these forecasting methods, and conclusions are provided in Chapter~\ref{ch6}.

@@

\section{Takens Embedding Theorem}\label{sect_Takens}

@@

Though not a uniquely related to the Takens embedding theorem, we do opt to make mention of this requirement for our work. 
The methods described here are applicable for data originating from a surjective map. 
This follows from the fact that when $T$ maps the space $U$ into a proper subset $B$ of $U$, we me conclude that $T$ is not surjective as there exists some point in $U-A$ which does not have a preimage.
An example of this would be the case when a system displays energy-loss. Consider, for instance, consider the oscillations dying down for a damped pendulum. (A more in-depth account will be given in \ref{ch5})

This requirement is not overly restrictive, as a great number of chaotic systems do have a surjective map. (Cite) \ednote{I rememeber you saying this in one of our sessions, but don't have a source for it.}

@@

\begin{Definition}
  [\bf Entire Solution] \label{Dfn_Soln} \rm
  A sequence $\{x_n\}_{n\in\mathbb{Z}}\subset X$ is called an entire solution (or simply a solution) of the driven system  $g$ with input $\overline{u}$ when it satisfies 
  \[g(u_{n-1}, x_{n-1})=x_n\] for all $n\in\mathbb{Z}$
\end{Definition}

@@

\begin{Definition}
  [\bf Unique Solution Property] \label{Dfn_usp}\rm
  A driven system $g$ is said to have the Unique Solution Property (USP) if for each input $\overline{u}$ there exists exactly one solution. 
  Alternatively we may formulate the USP as follows: $g$ has the Unique Solution Property if there exists a well-defined map $\Psi:{U}\to{X}$, with $\Psi({\overline{u}})$ denoting the unique solution.
\end{Definition}

One of the first result obtained after defining the USP is the fact that every solution will attract different initial conditions towards the component parts of the solution. %, i.e. that the solution $\Psi$ is a non-autonomous uniform attractor. \ednote{Define Uniform Attractor before stating theorem or else omit the statement.} 
If $g$ has the USP, then any solution to $g$ is also a uniform attractor in nonautonomous dynamical systems literature \cite{Manju_Nonlinearity}. The discussion on nonautonomous attractors is beyond the scope of this project and we refer the reader to \cite{Manju_ESP, esann2012ids} for further reading.

% \ednote{M: You can omit the theorem, and just say it is a uniform attractor in the nonautonomous dynamical systems literature \cite{Manju_Nonlinearity}, and say the discussion on nonautonomous attractors is beyond the scope of the thesis.}
% Paraphrased.
Additionally, $g$ having the USP is in our context also equivalent to $g$ being a topological contraction \cite{manjunath2021universal}, or alternatively also that $g$ exhibits the Echo State Property (ESP).

In the majority of Reservoir Computing(RC) literature (a machine learning algorithm approach)\ednote{How should I explain/qualify RC better here? It's mentioned without really explaining.}, 
the ESP is used to describe the concept of fading memory or memory-loss  \cite{jaeger2004harnessing, lu2018attractor}. 
If $g$ possesses the ESP (equivalent in our context to the USP), the we are guaranteed that the whole of an input's left-infinite history will exactly determine the current system state; i.e. there is only ever the possibility of a single reachable state at a given instance of time \cite{jaeger2001echo}

Defined formally, the ESP is intrinsically linked to an input (see \cite{Manju_ESP}) Unfortunately, definitions and results concerning the ESP do not have much practical benefit since a relationship between the network's response and the temporo-statistical properties of the input are not directly described. A more detailed presentation may be found in~\cite{jaeger2001echo}.

We do however consider it important to mention that the ESP is also often discussed in terms of a stability property that holds for all inputs of a system~\cite{manjunath2020stability} and, furthermore, plays a key role in the design and training of recurrent neural networks(RNN) within the field of RC \cite{Manju_ESP}.

@@

Having already defined the reachable set $X_U$ as the set of all elements of all entire solutions, we pause briefly in order to fix additional notation.
Defining $\cev{u}^{n}:=(\ldots,u_{n-2} ,u_{n-1})$ as the left-infinite subsequence of an input up until time $n$, $\overleftarrow{U}$ is then the notation for all these left-infinite sequences in $U$. 
Moreover, $\cev{u}^{n}v:=(\ldots,u_{n-2} ,u_{n-1}, v)$ will symbolise the input up to time $n$ with $v \in U$ being the specific input value at time $n$. 
The introduction of a new input at time $n$ can be described by the mapping $\sigma_v:   \cev{u}^{n} \mapsto \cev{u}^{n}v$.

The goal now assumes the form of the question on our ability to establish a semi-conjugacy as presented below for the driven system $g$. 
%\ednote{M: I have changed it to a semi-conjugacy since conjugacy amounts to asking a infinite-dimensional space $\overleftarrow{U}$ to be embedded in $X_U$ which is not possible.}

@@

We proceed to consider a specific subclass of conjugacies.

  \begin{Definition}
    [\bf Universal Semi-Conjugacy]\label{Def_UnivSemiConj} \rm
    Given a driven system $g$, we  call a continuous and surjective map $h : \overleftarrow{U} \to X_U$ a universal semi-conjugacy if  diagram \ref{Scomm_h} commutes for all $v \in U$.
  \end{Definition}

If the universal semi-conjugacy $h$ exists (i.e. the diagram in \ref{Scomm_h} commutes) then the solution $\Psi$ will be at least a coarse-grained representation of the input $u$. \ednote{B: Expand/Cite}

But does such a function $h$ for the above driven system $g$ above exist? Whenever $g$ has the USP and $\Psi(u)=\{x_n\}_{n\in\mathbb{Z}}$ it follows that $h$, defined by  $h(\cev{u}_n):=g(u_n,x_{n-1})=x_n$, will satisfy the semi-conjugacy in the graph above \ref{Scomm_h}.
Regrettably, such a continuous mapping $h$ is not guaranteed to exist \cite{M: Have edited here} when $g$ does not have the USP \cite[Lemma 5]{Manju_Nonlinearity}.
Note that even when $h$ does exist, we are not guaranteed its injectivity. Considering again example \ref{ex_halfux}, we see that even if $h$ were to exist, it could not be injective as $X_U=\{0\}$.

@@

\begin{Definition}
  [\bf Causal Mapping]\label{Def_CausMap}
  A continuous, surjective map $H:\overleftarrow{U}\to\overleftarrow{X}_U$ such that \[H\circ\tilde{g}_v=\sigma\circ{H}\] where $\tilde{g}_v$ maps $(\ldots, u_{-2}, u_{-1})$ to $(\ldots, u_{-2}, u_{-1}, g(v, u_{-1}))$ is called a causal mapping.
\end{Definition}

@@

 \begin{Theorem}
  For a compactly driven system, a causal mapping $H$ exists if and only if $g$ has the USP. 
\end{Theorem}
\begin{proof}
  See~\cite[Th.3]{manjunath2013echo}.
\end{proof}
To establish whether $\widehat{T}$ and $G_T$ are related, we consider the the function $H_2{\overline{}} := (h(r\overline{u}, h(\overline{u})))$. $H_2$ send the entirety of a left-infinite sequence to some element of $X\times{X}$.

% An embedding of the space $\overleftarrow{U}$  in $\overleftarrow{U}$ would allow one to establish topological conjugacy, which in turn provides stronger results than merely obtaining a coarse-grained representation via a causal mapping \ednote{M: I feel this is very confusing to the reader as the co-domain of causal mapping is the left-infinite space and , so please omit if possible. You can just say we want to restrict inputs to the left-infinite orbits of a dynamical system and then explore conjugacy instead of a mere semi-conjugacy}.

% Before formalising this, we make mention of the concept of the inverse-limit system of a dynamical system $(U,T)$. 
We would like to restrict inputs to the left-infinite orbits of a dynamical system and accordingly explore the possibility of establishing a conjugacy instead of merely obtaining a semi-conjugacy.
These spaces are noteworthy as their topology can help one ascertain much about the original system \cite{Manju_IEEE}. The topological structure for inverse-limit systems of chaotic dynamical systems is complicated, so we relegate further discussion to \cite{kennedy2008inverse_limit, ingram2011inverse}.
In broad sweeps, the inverse-limit space of a dynamical system is a self-map on a subset of an infinite-dimensional space where each point in the inverse-limit space corresponds to a backward orbit of the map $T$. 
Following \cite{manjunath2021universal, ingram2011inverse}:

@@

The driving function $g$ cannot be chosen in a frivolous manner as it could possibly complicate our work. Specifically, one must be careful to avoid a choice of $g$ which would  add complexity to the obtained solution.

@@

\begin{Definition}
  [\bf SI-Invertibility]\label{Dfn_SIinv}\rm
  A map $g$ is said to be SI-Invertible if $g$ is invertible for all $x\in{X}$. Alternatively it may be said that if, given $x_n$ and $x_{n-1}$, $u_{n-1}$ can be uniquely determined from $x_n=g(u_n,x_{n-1})$, then $g$ is said to be SI-invertible.
\end{Definition}

@@

\begin{equation}  \label{eqn_driving}
  g(u,x) = (1-a)x + a\overline{\tanh}(Au + \alpha Bx)
\end{equation} 
is both SI-invertible and, if $\alpha\beta<1$, also possesses the USP. 
One may easily show that $g$ is SI-invertible by recovering $u_n$ by 
\begin{equation} \label{eqn_SI_RNN}
  u_{n-1} := A^{-1}\bigg(\overline{\tanh}^{^{-1}}\frac{1}{a}\Big(x_{n+1}-(1-a)x_n\Big) \bigg) - \alpha B x_n
  \end{equation}
  for $x_{n-1}, x_n$ known.

@@

\begin{equation} \label{fig:inverse_limsystem}
  %  \[ 
      \psset{arrows=->, arrowinset=0.25, linewidth=0.6pt, nodesep=3pt, labelsep=2pt, rowsep=0.7cm, colsep = 1.1cm, shortput =tablr}
      \everypsbox{\scriptstyle}
      \begin{psmatrix}
      \widehat{U}_T  && \widehat{U}_T \\
      Y_T && Y_T \\
      %%%
     %  \ncline{1,1}{1,2}^{\widehat{T}} \ncline{1,1}{2,1} <{h}
     %  \ncline{1,2}{2,2} > {H_2}
     %  \ncline{2,1}{2,2}^{G_T}
      \end{psmatrix}
    %  \]
  \end{equation}

@@

The graph in equation \ref{fig:inverse_limsystem} is then extended as below:

@@

From \eqref{eqn_driving} we know that $ g(u,x)=(1-a)x+a{\cdot}\tanh(Au+\alpha{a}Bx)$. Recall that $g$ is SI-invertible. Adopting the terminology of \cite{manjunath2021universal}, $A$ is called the “input matrix” and $B$ the “reservoir matrix”.

We make use of a Recurrent Neural Network (RNN) in our implementation. RNN's are widely used in the literature for discrete-time processing and are considered to be discrete-time version of continuous dynamical systems modelling a network of neurons. (\textbf{Citation.})
A complete discussion of RNN's is outside the scope of this thesis, so we present only the details necessary to the current discussion. For information, (\textbf{Cite, Cite, Cite}) may be consulted. We implement our RNN in the \emph{Python} programming language and making use of the \emph{Scipy} library, alongside the \emph{Keras} package in the \emph{Tensorflow} library.

@@

To understand the first and most drastic\ednote{change word} change, we recall the the conjugacy in~\ref{fig:inverse_limsystem}:

\begin{equation*}
  %  \[ 
      \psset{arrows=->, arrowinset=0.25, linewidth=0.6pt, nodesep=3pt, labelsep=2pt, rowsep=0.7cm, colsep = 1.1cm, shortput =tablr}
      \everypsbox{\scriptstyle}
      \begin{psmatrix}
      \widehat{U}_T  && \widehat{U}_T \\
      Y_T && Y_T \\
      %%%
     %  \ncline{1,1}{1,2}^{\widehat{T}} \ncline{1,1}{2,1} <{h}
     %  \ncline{1,2}{2,2} > {H_2}
     %  \ncline{2,1}{2,2}^{G_T}
      \end{psmatrix}
    %  \]
  \end{equation*}

  We recall that $G_T$ is topologically conjugate to $\widehat{T}$, a result which we have already made use of.

  The authors of~\cite{manjunath2021universal} did not, however, consider the relationship between the inverse-limit space $\widehat{U}_T$ and the input $u_n=\theta(w_n)$ (where $\theta$ is our observable and $w_n$ the complete input), a concept we now explore.
  We may construct a delay-coordinate $v_n=[\theta(w_n), \theta(w_{n-1}), \ldots, \theta(w_{n-d})]^{T} = [u_n, u_{n-1}, \ldots, u_{n-d}]^T$
    to accordingly obtain a new dynamical system $(V,F)$ where $F$ is the map $F:v_n\mapsto{v_{n+1}}$. If $d$ is sufficiently large, we may apply Takens Embedding Theorem~\ref{Thm_Takens} to establish the existence of a topological conjugacy between the new system $(V,F)$ and the original system $(U,T)$.

  Using the same notation as before, we may construct a new conjugacy which now relates delay coordinates to one another.

  \begin{equation}\label{eqn_VFdelay}
    %  \[ 
        \psset{arrows=->, arrowinset=0.25, linewidth=0.6pt, nodesep=3pt, labelsep=2pt, rowsep=0.7cm, colsep = 1.1cm, shortput =tablr}
        \everypsbox{\scriptstyle}
        \begin{psmatrix}
        \widehat{V}_F  && \widehat{V}_F \\
        Y_F && Y_F \\
        %%%
       %  \ncline{1,1}{1,2}^{\widehat{F}} \ncline{1,1}{2,1} <{h}
       %  \ncline{1,2}{2,2} > {*}
       %  \ncline{2,1}{2,2}^{G_F}
        \end{psmatrix}
      %  \]
    \end{equation}

The existence of $\widehat{F}$ is guaranteed, but the map $v_{n}\mapsto{v_{n+1}}$ itself is unknown.
We may, however, make use of the SI-invertibility of the function $g(u_{n-1}, x_{n-1})=x_n$ to conclude that
$(x_{n-1}, x_{n})$ determines $u_{n-1}$ uniquely.

Accordingly $(x_{n-2}, x_{n-1}, x_{n})$ determines $u_{n-2}$ and $u_{n-1}$ uniquely and generally $(x_{n-d}, \ldots , x_{n-1}, x_{n})$ determines $u_{n-d}, \ldots u_{n-2}$ and $u_{n-1}$ uniquely.

We therefore need not know what $\widehat{F}$ is, as we can determine $v_{n+1}={[u_n, u_{n-1}, \ldots, u_{n-d}]}^T$ indirectly by determining the value of $(x_{n-d}, \ldots , x_{n-1}, x_{n})$, a much more feasible task.

Accordingly we define the following abbreviations:
\begin{itemize}
  \item Delay Coordinates (DC): A vector composed by appending a fixed amount of previous values to the final value.
  \item Single-Delay Dynamics(SDD): Dynamics generated by 2-element delay-coordinate vectors of the form $(x_{n-1}, x_n)$ where only one previous value is used.
  \item Multi-Delay Dynamics(MDD): Dynamics generated by arbitrary-length delay-coordinate vectors of the form $(x_{n-d}, \ldots , x_{n-1}, x_{n})$. ($d$ must be sufficiently large, but is unknown. We vary this parameter in experimentation.)
\end{itemize}

Within this framework, the PCA-implementation of~\cite{manjunath2021universal} is then considered to fall underneath the of auspices of what we term \emph{Approximated Multi-Delay Dynamics(AMDD)}.

We have now seen that one may comfortably switch between the use of SDD or MDD when $g$ has SI-invertibility. /ednote{This sentence needs to be improved.}

Naturally, one would immediately suppose that the MDD approach would increased computation time and resource use as the internal matrices and datasets generated during learning would grow with every increase in the delay-parameter $d$.

In fact, preliminary results indicate the opposite due the observation that the implementation of MDD is inversely proportional to the the dimension of the network (i.e. the number of neurons in the RNN) and possibly also the number of epochs/batch size necessary to produce reliable predictions. 
This, in turn, actually speeds up the computation-time and provides results of similar accuracy. \ednote{I did not have enough time to explore this properly. I can only hazard guesses and promise future research work in this project. How should I best convey this? }

\textbf{Secondly}

Sparse Matrices:
\vspace{-8mm}
\begin{itemize}[noitemsep]
\item Well-conditioned: condition number close to 1
\item Scaled
\item Sparsity/density ratio
\end{itemize}

% Sparse matrix implementation. Slight improvement in some scenarios, but not fantastic. \ednote{M: Did it not help you a great deal with the DP forecasting? B: Sparse matrices tended to be finicky. In some cases they helped whereas in others they actually worsened the solution somewhat. There's some tuning to be done in terms of condition number, scaling and density}

\textbf{Third}

Density of matrices A/B(i.e. number of non-zero entries in each matrix) (separate from sparse matrix implementation)

@@

The scalar equations of motion are derived using Newtonian physics and can be found in~\cite{DPFormulas}. The second-order system is converted to a first-order ODE and solved using the Runge-Kutta numerical method.  Noise-free, undamped DP data-sets were generated and utilised in this project due to the requirement that the map $T$ be surjective. We discuss the case of a DP exhibiting energy-loss (i.e. where the map $T$ is not surjective) below.

\textbf{Data originating from a non-surjective map.}

The importance of the requirement that data originates from a surjective map is observed learning failure of a map describing a damped DP.
Due to the assumption, the program discounted any possibility of energy-loss and we consider the example below.

\begin{figure}[ht]
  \includegraphics[scale=0.4]{_dp_dying.eps}
  \centering
\caption{The map generating data for a damped pendulum is not surjective.}
\label{fig:damped_pendulum}
\end{figure}

Despite feeding data that originated from a DP with dying occilations into the system, a map was instead learnt which continued to oscillate in a chaotic, but conservative manner.

\begin{figure}[ht]
  \includegraphics[scale=0.4]{_dpfailure_nonsurj.eps}
  \centering
\caption{Failure to learn a map which is not surjective.}
\label{fig:dp_notsurjective}
\end{figure}

Having spent ample time to discuss the challenges, we now change tone so as to satisfy the optimist.
We present successful run for DP-data to repdict the future movement of the pendulum.

The system was driven for 15500 steps, of which the first 500 were immediately discarded to simulate the network 'forgetting'. The driving function $g$ had input and reservoir matrices were implemented as sparse matrices filled to 20\%. Parameter-values of $\alpha=0.99$ and $a=0.5$ were used.
The remaining 15000 datapoints were used to train a network with 128 neurons. 
The network was initialised to contain 16 hidden layers each with a layer dimension of 64, activated with the tanh activation function.
Training was accomplished using 256 epochs with batch sizes 128 and a delay of 1.

% Trainlen:15000; Predlen: 1000; Discard: 500; Dimension:128 neurons; $\alpha=0.99$; $a=0.5$. Hidden Layers: 16. Layer dimension: 64. Epochs:256. Batch Size: 128. Delay(for MDD): 8. Density of Sparse Matrices: 0.2. $d=1$

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.8\linewidth]{_DP_Fig-2F2023.eps} 
  \captionof*{figure}{Predicted(red) and actual(blue) trajectories of the $x$- and $y$-coordinates of each of the pendulum heads.} 
% \end{figure}
% \begin{figure}[ht]
%   \centering
  \includegraphics[width=0.8\linewidth]{_DP_Fig-2B2023.eps}
  \captionof*{figure}{Densities of predicted and actual values of x-coordinate of first pendulum head. }
  % \end{figure}
  % \begin{figure}[ht]
  %   \centering 
  \includegraphics[width=0.8\linewidth]{_DP_Fig-2D_5.eps}
    \captionof*{figure}{Takens Delay Map with delay=5 to illustrate behaviour of predicted and actual values of double pendulum data.}
    \label{fig:dp_predictions}
\end{figure}

\subsection{Additional Attractors: Clifford \& Thomas}
We follow the methodology presented in \cite{manjunath2021universal} but opt to consider attractors different from the Lorenz system and H\'enon map in this project.

@@

The behaviour of the system with a $\beta$-parameter value of 0.1056 was simulated by scaling it to fit inside the interval $[-1,1]$. Two cases were considered: noise-free and perturbed. Noise was generated from a normal distribution with mean 0 and standard deviation 0.05, which translates approximately to *dB.
 A sequence of 3000 observations in both the clear and noisy data-sets respectively were fed into the system with 300 entries discarded to simulate the network's loss of memory. The next 9000 steps were predicted and compared with true data.

@@

The RNN was initialised as follows: the dimension of the network is set equal to 500 neurons and 12 hidden layers of dimension 64 each were initialised. Training is realised through 150 epochs of batch sizes of 128 each; the map is learnt via the Adam Optimiser using finer and finer learning rates (0.001 and then divided by 10 at every iteration).

@@
